The influence of Gaia-Sausage-Enceladus-like merger events on the chemical evolution of galaxy discs

### 2023-03-09 src/tex/ms.tex
[showyourwork] overleaf sync

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title[Early massive accretion events in MW-mass galaxies]{The influence of Gaia-Sausage-Enceladus-like merger events on the chemo-dynamics of galaxy discs}
%The influence of massive early merger events on the chemo-dynamics of galaxy discs

\author[T. Buck] {Tobias Buck$^{1,2}$\thanks{E-mail: tobias.buck@iwr-uni-heidelberg.de.de}, Aura Obreja$^{3}$, \etal \\  %Andrea V. Macci\`o$^{4,5}$, \etal \\
%, Aaron A. Dutton$^{2}$, Andrea V. Macci\`o$^{2,1}$\\
%
$^1$Universit\"at Heidelberg, Interdisziplin\"ares Zentrum f\"ur Wissenschaftliches Rechnen, Im Neuenheimer Feld 205, D-69120 Heidelberg, Germany\\
$^2$Universit\"at Heidelberg, Zentrum f\"ur Astronomie, Institut f\"ur Theoretische Astrophysik, Albert-Ueberle-Straße 2, 69120 Heidelberg, Germany\\
$^3$Universit\"ats-Sternwarte M\"unchen, Scheinerstraße 1, D-81679 M\"unchen, Germany%\\
%$^4$New York University Abu Dhabi, PO Box 129188, Saadiyat Island, Abu Dhabi, United Arab Emirates\\
%$^5$Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, 69117 Heidelberg, Germany\\
}

@@

\begin{abstract}
awesome science! keep on reading and cite me! 
\end{abstract}

@@

Our Milky Way (MW) Galaxy is the best-studied galaxy in the
universe and provides the most stringent constraints of galaxy formation
models \citep[e.g.,][]{Guedes11, Wetzel16, Grand17, Buck21}. Therefore, a thorough understanding of the components of the Galaxy, their dynamical evolution and formation channel is of great interest.

The Galaxy's main body, the rotationally-supported stellar disk harbors multiple components or populations when dissected in the space of chemistry, kinematics, spatial extent, and age \citep[e.g.,][]{Gilmore83, Norris85, Chiba00, Nissen10, Bovy12, Haywood13}; see \citet{Rix13} and \citet{Bland-Hawthorn16} for recent reviews. %There exist a dichotomy in both spatial structure and stellar chemistry. Spatially, MW's disk can be separated into a thin and a thick disk with different scale heights. Chemically, a clear bimodality in the abundance of $\alpha$ elements relative to iron ([$\alpha$/Fe]) has been found. In general, the high$-\alpha$ population has older ages and a larger scale-height than the low$-\alpha$ population. However, the mapping between the high$-\alpha$ (low$-\alpha$) and the thick (thin) disks is not perfect.

Thanks to the great improvements in the quality and volume of astrometric datasets provided by the {\it Gaia} mission in combination with chemical abundances and radial velocities from large spectroscopic surveys, (e.g. \textsc{apogee}; \textsc{galah}; \textsc{h3}; \textsc{LAMOST}; \citealt{Majewski2017}, \citealt{Martell2017}, \citealt{Conroy2019}, \citealt{Zhao_2012}, respectively) we are now able to identify and link coherent chemodynamical structures in the solar neighbourhood to ancient merger events several billion years back in MW's history (e.g. \citealt{Helmi2020} for a review). By now a large number of chemocynamical clusters associated with accreted systems have been identified in the high-dimensional space of orbital parameters and stellar abundances \citep[e.g.,][]{Belokurov18, Helmi18, Myeong19a, Naidu20a, Horta21, Myeong2022}.

Most notably, there exists an excess of stars on radial orbits in the local stellar halo around the Sun which is often referred to as {\it Gaia}-Sausage-Enceladus (GSE; \citealt{Belokurov2018, Helmi2018}; see also \citealt{Nissen2010, Koppelman2018, Haywood2018}). This stellar overdensity has been identified as the observable signature of a massive ($M_{\star} \simeq 10^{9}$ $\rm{M_{\odot}}$) early ($8-11$ Gyr ago; e.g. \citealt{Vin2019, Bel_2020, Naidu21, Xiang22}) accretion event. GSE is by far the most significant merger in MW's history and contributed approximately two-thirds of MW's stellar halo stars on highly-eccentric orbits \citep[e.g.][]{Mackereth+Bovy20}. GSE stars inhabit a sausage-like distribution in the radial-azimuthal velocity distribution \citep{Brook2003, Belokurov2018} and appear to be more metal-poor and less $\alpha$-enhanced than the redder halo counterpart \citep{Haywood_2018a, Helmi_2018}. The redder sequence of the Galactic halo is thought to be the result of proto-galactic disc stars being dynamically ejected into the halo during the GSE merger and has been associated with an event termed {\it the Splash} by \cite{Belokurov2020, Bonaca2020}, initially also found by other earlier studies \citep{Bonaca2017, Haywood2018, DiMatteo2019, Gallart2019}.

The existence of the GSE merger and its strong impact on the structure of the proto-MW demands a thorough study of the processes that shape MW's chemo-dynamical evolution during such a merger in order to interpret the observational data at hand. In order to connect present day observables to Gyr old events we need tools like cosmological simulations that model galactic mass growth, mergers and subsequent star formation plus chemical enrichment self-consistently. Here, enough resolution to follow the internal disk dynamics and suppress spurious heating ($N \gtrsim 10^6$; e.g. \citealt{Sellwood2013,Ludlow2019DiskHeating, Ludlow2021}) as well as resolving the multi-phase, dense structure of the interstellar medium (ISM) combined with a proper model for chemical enrichment \citep[e.g.][]{Buck2021} is key to accurately capture their birth kinematics and the subsequent dynamical evolution of the disk (e.g. \citealt{House2011, Bird2013}).

Thanks to recent advances in computing power and progress in numerical methods modern cosmological zoom simulations now meet the aforementioned requirements and are able to model individual MW-mass galaxies by sampling varying environments and formation scenarios (e.g. \citealt{Sawala2016, Grand2017, Buck2020, Font2020, Applebaum2021, Agertz2020Vintergatan, Bird2021, Khoperskov2022-1InSitu, Wetzel2022}). This enables us to reconstruct our Galaxy's formation history by linking the occurrence of chemodynamical patterns at $z=0$ to specific events in the Galaxy's (e.g. \citealt{Bignone2019, Mackereth2019, Fattahi2019, Grand2020, Elias2020, Dillamore2022, Khoperskov2022-2Accreted, Khoperskov2022-3Metallicity, Pagnini2022, Rey2022}).

While GSE's strong impact on MW's stellar halo is by now well established, more recently studies have further investigated its defining influence on the formation of MW's thin and thick disk \citep[e.g.]{Grand2020,Ciuca2022,Orkney2022,Rey2023}. These studies suggest that the GSE merger strongly shaped the formation of MW's thick disk agreeing with thick disk formation scenarios from early simulation results by \cite{Brook_2004, Brook_2006}. The gas-rich merger has a two-fold effect on MW's stellar disk: (i) Firstly, it heats part of the existing proto-disc stars dynamically ejecting them into the halo and creating the Splash. (ii) Secondly, the merger provides fresh gas to the central galactic regions triggering a starburst that eventually forms the younger thick disc. After the merge, the thin disc then forms from the accretion of metal-poor gas in an inside-out, upside-down fashion \citep[e.g.,][]{Bird2013, Grand2018, Buck2020a}. \citet{Ciuca_2021} qualitatively confirmed this picture using APOGEE DR14 data and selecting stellar populations around the solar radius.

Interestingly, when inferring the birth radii of MW disk stars, \citet{Lu2021} found a steepening of the metallicity gradient at the time of the GSE merger using data from LAMOST combined with {\it Gaia} eDR3. Here we use the NIHAO-UHD suite \citep{Buck2020} of cosmological hydrodynamical simulations of MW-mass galaxies in order to investigate the influence of early merger events on the chemo-dynamics of the stellar disk. By using 4 different simulations from the NIHAO-UHD suite we investigate independently sampled environments that follow 4 different formation scenarios. Especially we focus on the frequency of massive early mergers and their detailed impact on the formation of the stellar disk. This paper is structures as follows: In \S 2 we describe the simulations on which we base our analysis, followed by a presentation of our results in \S 3. In \S 4 we discuss and conclude our findings.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results} \label{sec:results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{The impact of gas rich merger on the metallicity gradient} \label{sec:merger}
We start our analysis on the influence of Gaia-Sausage-Enceladus-like merger
events on the chemo-dynamics of MW-like galaxy disks by looking at the merger history of the four galaxies. Figure~\ref{fig:merger_ratio} shows the cumulative gas mass growth (blue line, left axis) accompanied by the gaseous merger ratio (orange line, right axis) defined as the ratio of gas mass of the merging satellite over the gas mass of the main galaxy. We see that after an initial phase ($0-\sim2.5$ Gyr) of rapid gas mass growth accompanied by violent, relatively gas rich mergers there are several other mergers at later times that contribute more than 10\% in gas leading to sudden jumps in the gas mass of the main galaxy (blue line). 
Especially for g2.79e12 (left most panel) and g7.08e11 (right most panel) we see some late time ($\sim10$ Gyr) gas rich merger while the other two galaxies do not show any gas rich merger after $5$ Gyr for g8.26e11 and $7$ Gyr for g7.55e11.

@@

\begin{figure}
    \script{gas_profile_all.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/gas_profile_2d.pdf}
        \caption{
            Relative change in cold gas surface density profile before and after the merger events as indicated in Fig.~\ref{fig:merger_ratio} by the gray lines. Outside of a radius of $\sim5-10$ kpc the merging satellites contribute significant amounts of cold gas up to 200\%.
        }
        \label{fig:surf_den}
    \end{centering}
\end{figure}

@@

@@

In order to investigate the cause of the strong steepening of the metallicity gradient in our simulations we first plot in Figure~\ref{fig:surf_den} where in the disk the accreted gas mostly contributes. Figure~\ref{fig:surf_den} plots the relative change of cold gas surface mass density radial profile before the gas rich merger and after the merger finished as indicated by the gray lines in Figure~\ref{fig:half_mass}. This figure reveals an interesting finding. Consistently across all four galaxies the surface mass density of cold gas in the inner $5$ kpc is roughly constant or decreases slightly due to gas consumption by star formation and gas expulsion by feedback. Outside of $5$ kpc the surface mass density increases by $\sim75$ \% up to $\sim200$\% due to cold gas added byt he merger event, either by direct accretion of cold gas from the merging satellite or by triggered gas accretion from the circum-galactic medium due to tidal forces exerted by the interacting satellite.

Three scenarios can now lead to a steepenig of the metallicity gradient: (i) the metallicity of the star forming gas in the outskirts of the galaxies is diluted by the freshly accreted gas; (ii) the central parts of the galaxy keep enriching more strongly than the outskirts or (iii) a mix of both.
In fact, as was shown in \citet{Buck2020} in their Fig. 8 scenario (iii) is happening in these galaxies. For most of the time the ISM/cold gas metallicity evolution is self-similar and almost monotonically increasing at each radius but with markedly dilution events that correlate with merger events. In order to investigate this in more depth and study its effect on the metallicity gradient, we plot in Fig.~\ref{fig:feh_evolution} the difference in metallicity of the cold gas, $\Delta\mathrm{[Fe/H]}=\mathrm{[Fe/H]}(R<2\, \mathrm{kpc})-\mathrm{[Fe/H]}(R)$, between the gas in a given annulus at a radius $R$ and the metallicity in the central parts ($R<2$ kpc) of the galaxy, $\mathrm{[Fe/H]}(R<2\, \mathrm{kpc})$ as a function of time. We normalize this difference in metallicity, $\Delta\mathrm{[Fe/H]}$, by the central metallicity in order to take out the continued enrichment in [Fe/H] and highlight deviations from the self-similar evolution which is indicating a steepening or flattening of the metallicity gradient as measured over different radial ranges. Thus, a flat line in Fig.~\ref{fig:feh_evolution} would indicate perfect self-similar evolution while a more negative value indicates that a certain radius lacks behind the central galaxy in [Fe/H] enrichment and thus a steepening of the metallicity gradient. On the other hand, a progressively more positive or less negative value shows that the radius at hand enriches more strongly in metallicity than the central parts an thus indicates a flattening of the metallicity gradient.

\textcolor{red}{make same figure for relative metallicity change or similar to buck2020 Fig. 8 metallicity in different radii but normalized to central bin}

@@

\begin{figure*}
    \script{enrichment_evolution.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/enrichment_evolution.pdf}
        \caption{
            some caption
        }
        \label{fig:feh_evolution}
    \end{centering}
\end{figure*}

@@

\subsection{The impact on the age-metallicity relation and the [O/Fe] vs. [Fe/H] plane}
\label{sec:g2.79e12}

We now turn to investigate the question where the early gas rich mergers appear in the age metallicity relation and the [O/Fe] vs. [Fe/H] plane. To this extent we focus on the galaxy g2.79e12 for which the steepening of the metallicity gradient is strongest. Figure~\ref{fig:mdf} shows the metallicity distribution function (MDF) for gas (orange histogram) and stars (blue histogram) of three gas rich satellites that merge with the main galaxy at $2.02$ Gyr and $2.57$ Gyr in comparison to the MDF of gas (gray histogram) and stars (black histogram) in the central galaxy.

\begin{figure*}
    \script{mdf.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/2.79e12_mdf_gas.pdf}
        \caption{
            Metallicity distribution function of the main galaxy and the merging satellite shortly before coalescence for all three merger events. Filled histograms show the gaseous MDF while steps show the MDF for stars. In any case, the gas metallicity of the satellite is $\sim0.5-0.75$ dex lower that the main galaxy's gas metallicity.
        }
        \label{fig:mdf}
    \end{centering}
\end{figure*}

\begin{figure*}
    \script{age_feh.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/2.79e12_age_metallicity_grid.pdf}
        \caption{
            Age-metallicity relation.
        }
        \label{fig:age_feh}
    \end{centering}
\end{figure*}

\begin{figure*}
    \script{feh_ofe.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/2.79e12_feh_ofe_grid.pdf}
        \caption{
            Oxygen vs. metallicity.
        }
        \label{fig:ofe_feh}
    \end{centering}
\end{figure*}

@@

Our results are summarized as follows:
\begin{itemize}
%
\item

\end{itemize}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title[Early massive accretion events in MW-mass galaxies]{The influence of Gaia-Sausage-Enceladus-like merger events on the chemical evolution of galaxy discs}
%
%{The influence of Gaia-Sausage-Enceladus-like merger events on the chemo-dynamics of galaxy discs}
%The influence of massive early merger events on the chemo-dynamics of galaxy discs

\author[T. Buck] {Tobias Buck$^{1,2}$\thanks{E-mail: tobias.buck@iwr-uni-heidelberg.de.de}, Aura Obreja$^{3}$, \etal \\  %Andrea V. Macci\`o$^{4,5}$, \etal \\
%, Aaron A. Dutton$^{2}$, Andrea V. Macci\`o$^{2,1}$\\
%
$^1$Universit\"at Heidelberg, Interdisziplin\"ares Zentrum f\"ur Wissenschaftliches Rechnen, Im Neuenheimer Feld 205, \\ D-69120 Heidelberg, Germany\\
$^2$Universit\"at Heidelberg, Zentrum f\"ur Astronomie, Institut f\"ur Theoretische Astrophysik, Albert-Ueberle-Straße 2, \\ D-69120 Heidelberg, Germany\\
$^3$Universit\"ats-Sternwarte M\"unchen, Scheinerstraße 1, D-81679 M\"unchen, Germany%\\
%$^4$New York University Abu Dhabi, PO Box 129188, Saadiyat Island, Abu Dhabi, United Arab Emirates\\
%$^5$Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, 69117 Heidelberg, Germany\\
}

@@

\begin{abstract}
awesome science! \\
keep on reading \\
and cite me, please!\\
we have really great results!\\
the greatest results you will see today...\\
I hope this is enough to fill some lines \\
and add some space that might be filled by an actual abstract.\\
Let's see \\

\end{abstract}

@@

Our Milky Way (MW) Galaxy is the best-studied galaxy in the
universe and provides the most stringent constraints of galaxy formation
models \citep[e.g.,][]{Guedes2011, Wetzel2016, Grand2017, Buck2020a}. Therefore, a thorough understanding of the components of the Galaxy, their dynamical evolution and formation channel is of great interest.

The Galaxy's main body, the rotationally-supported stellar disk harbors multiple components or populations when dissected in the space of chemistry, kinematics, spatial extent, and age \citep[e.g.,][]{Gilmore1983, Norris1985, Chiba2000, Nissen2010, Bovy2012, Haywood2013}; see \citet{Rix2013} and \citet{Bland-Hawthorn2016} for recent reviews. %There exist a dichotomy in both spatial structure and stellar chemistry. Spatially, MW's disk can be separated into a thin and a thick disk with different scale heights. Chemically, a clear bimodality in the abundance of $\alpha$ elements relative to iron ([$\alpha$/Fe]) has been found. In general, the high$-\alpha$ population has older ages and a larger scale-height than the low$-\alpha$ population. However, the mapping between the high$-\alpha$ (low$-\alpha$) and the thick (thin) disks is not perfect.

Thanks to the great improvements in the quality and volume of astrometric datasets provided by the {\it Gaia} mission in combination with chemical abundances and radial velocities from large spectroscopic surveys, (e.g. \textsc{apogee}; \textsc{galah}; \textsc{h3}; \textsc{LAMOST}; \citealt{Majewski2017}, \citealt{Martell2017,Buder2021}, \citealt{Conroy2019}, \citealt{Zhao2012}, respectively) we are now able to identify and link coherent chemodynamical structures in the solar neighbourhood to ancient merger events several billion years back in MW's history (e.g. \citealt{Helmi2020} for a review). By now a large number of chemodynamical clusters associated either with distinct in-situ components at the time of birth of the MW's stellar disk \citep[e.g.][]{Belokurov2022,Conroy2022} or associated with distinct accreted systems have been identified in the high-dimensional space of orbital parameters and stellar abundances \citep[e.g.,][]{Myeong2019, Naidu2020, Horta2021, Buder2022}.

Most notably, there exists an excess of stars on radial orbits in the local stellar halo around the Sun which is often referred to as {\it Gaia}-Sausage-Enceladus (GSE; \citealt{Belokurov2018, Helmi2018}; see also \citealt{Nissen2010, Koppelman2018, Haywood2018}). This stellar overdensity has been identified as the observable signature of a massive ($M_{\star} \simeq 10^{9}$ $\rm{M_{\odot}}$) early ($8-11$ Gyr ago; e.g. \citealt{Vincenzo2019, Belokurov2020, Naidu2021, Xiang2022}) accretion event. GSE is by far the most significant merger in MW's history and contributed approximately two-thirds of MW's stellar halo stars on highly-eccentric orbits \citep[e.g.][]{Mackereth2020}. GSE stars inhabit a sausage-like distribution in the radial-azimuthal velocity distribution \citep{Brook2003, Belokurov2018} and appear to be more metal-poor and less $\alpha$-enhanced than the redder halo counterpart \citep{Haywood2018, Helmi2018}. The redder sequence of the Galactic halo is thought to be the result of proto-galactic disc stars being dynamically ejected into the halo during the GSE merger and has been associated with an event termed {\it the Splash} by \cite{Belokurov2020, Bonaca2020} which has been initially found by other earlier studies \citep{Bonaca2017, Haywood2018, DiMatteo2019, Gallart2019}.

The existence of the GSE merger and its strong impact on the structure of the proto-MW demands a thorough study of the processes that shape MW's chemo-dynamical evolution during such a merger in order to interpret the observational data at hand. In order to connect present day observables to Gyr old events we need tools like cosmological simulations that model galactic mass growth, mergers and subsequent star formation plus chemical enrichment self-consistently. Here, enough resolution to follow the internal disk dynamics and suppress spurious heating ($N \gtrsim 10^6$; e.g. \citealt{Sellwood2013,Ludlow2019,Ludlow2021}) as well as resolving the multi-phase, dense structure of the interstellar medium (ISM) combined with a proper model for chemical enrichment \citep[e.g.][]{Buck2021} is key to accurately capture their birth kinematics (e.g. \citealt{Ratcliffe2022, Lu2022}) and the subsequent dynamical evolution of the disk (e.g. \citealt{House2011, Bird2013, Buck2019b}).

@@

Thanks to recent advances in computing power and progress in numerical methods modern cosmological zoom simulations now meet the aforementioned requirements and are able to model individual MW-mass galaxies by sampling varying environments and formation scenarios (e.g. \citealt{Sawala2016, Grand2017, Buck2018, Font2020, Applebaum2021, Agertz2021, Bird2021, Khoperskov2022-InSitu}). This enables us to reconstruct our Galaxy's formation history by linking the occurrence of chemodynamical patterns at $z=0$ to specific events in the Galaxy's \citep[e.g.][]{Bignone2019, Fattahi2019, Grand2020, Elias2020, Dillamore2022, Khoperskov2022a, Rey2022}.

While GSE's strong impact on MW's stellar halo is by now well established, more recently studies have further investigated its defining influence on the formation of MW's thin and thick disk \citep[e.g.]{Grand2020,Ciuca2022,Orkney2022,Rey2023}. These studies suggest that the GSE merger strongly shaped the formation of MW's thick disk agreeing with thick disk formation scenarios from early simulation results by \cite{Brook2004, Brook2006}. The gas-rich merger has a two-fold effect on MW's stellar disk: (i) Firstly, it heats part of the existing proto-disc stars dynamically ejecting them into the halo and creating the Splash. (ii) Secondly, the merger provides fresh gas to the central galactic regions triggering a starburst that eventually forms the younger thick disc. After the merge, the thin disc then forms from the accretion of metal-poor gas in an inside-out, upside-down fashion \citep[e.g.,][]{Bird2013, Grand2018, Buck2020a}. \citet{Ciuca2021} qualitatively confirmed this picture using APOGEE DR14 data and selecting stellar populations around the solar radius.

Interestingly, when inferring the birth radii of MW disk stars, \citet{Lu2022b} found a steepening of the metallicity gradient at the time of the GSE merger using data from LAMOST combined with {\it Gaia} eDR3. Here we use the NIHAO-UHD suite \citep{Buck2020a} of cosmological hydrodynamical simulations of MW-mass galaxies in order to investigate the influence of early merger events on the chemo-dynamics of the stellar disk. By using 4 different simulations from the NIHAO-UHD suite we investigate independently sampled environments that follow 4 different formation scenarios. Especially we focus on the frequency of massive early mergers and their detailed impact on the formation of the stellar disk. This paper is structures as follows: In \S~2 we describe the simulations on which we base our analysis, followed by a presentation of the impact of early massive mergers on the metallicity gradient of the cold gas disk in \S~3. In \S~4 we turn to analyse where merger stars and gas end up in the age-metallicity relation  and the [$\alpha$/Fe] vs. [Fe/H] plane of the main galaxy. In \S~5 we summarize and conclude our findings.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Impact on the cold gas metallicity gradient} \label{sec:results-gradient}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We start our analysis on the influence of Gaia-Sausage-Enceladus-like merger
events on the chemo-dynamics of MW-like galaxy disks by looking at the merger history of the four galaxies. Figure~\ref{fig:merger_ratio} shows the cumulative gas mass growth (blue line, left axis) accompanied by the gaseous merger ratio (orange line, right axis) defined as the ratio of gas mass of the merging satellite over the gas mass of the main galaxy. We see that after an initial phase ($0-\sim2.5$ Gyr) of rapid gas mass growth accompanied by violent, relatively gas rich mergers there are several other mergers at later times that contribute more than 10\% in gas leading to sudden jumps in the gas mass of the main galaxy (blue line). 
Especially for g2.79e12 (left most panel) and g7.08e11 (right most panel) we see some late time ($\sim10$ Gyr) gas rich merger while the other two galaxies do not show any gas rich merger after $5$ Gyr for g8.26e11 and $7$ Gyr for g7.55e11.

\subsection{The impact of gas rich merger on the metallicity gradient} \label{sec:merger}

@@

\begin{figure}
    \script{gas_profile_all.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/gas_profile_2d.pdf}
        \vspace*{-1.75em}
        \caption{
            Relative change in cold gas surface density profile before and after the merger events as indicated in Fig.~\ref{fig:merger_ratio} by the gray lines. Outside of a radius of $\sim5-10$ kpc the merging satellites contribute significant amounts of cold gas up to 200\%.
        }
        \label{fig:surf_den}
    \end{centering}
\end{figure}

@@

\begin{figure*}
    \script{enrichment_evolution.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/enrichment_evolution.pdf}
        \vspace*{-1.75em}
        \caption{
            Cold gas metallicity difference $\Delta\mathrm{[Fe/H]}=\mathrm{[Fe/H]}(R<2\, \mathrm{kpc})-\mathrm{[Fe/H]}(R)$, between the gas in a given annulus at a radius $R$ (colored lines) and the metallicity in the central parts ($R<2$ kpc) of the galaxy, $\mathrm{[Fe/H]}(R<2\, \mathrm{kpc})$ as a function of time. A flat line indicates perfect self-similar evolution between two different radii while a more negative value indicates that a certain radius lacks behind the central galaxy in [Fe/H] enrichment and is thus indicative of a steepening of the metallicity gradient.
        }
        \label{fig:feh_evolution}
    \end{centering}
\end{figure*}

@@

In order to investigate the cause of the strong steepening of the metallicity gradient in our simulations we first plot in Figure~\ref{fig:surf_den} where in the disk the accreted gas mostly contributes. Figure~\ref{fig:surf_den} plots the relative change of cold gas surface mass density radial profile before the gas rich merger and after the merger finished as indicated by the gray lines in Figure~\ref{fig:half_mass}. This figure reveals an interesting finding. Consistently across all four galaxies the surface mass density of cold gas in the inner $5$ kpc is roughly constant or decreases slightly due to gas consumption by star formation and gas expulsion by feedback. Outside of $5$ kpc the surface mass density increases by $\sim75$ \% up to $\sim200$\% due to cold gas added byt he merger event, either by direct accretion of cold gas from the merging satellite or by triggered gas accretion from the circum-galactic medium (CGM) due to tidal forces exerted by the interacting satellite.

Three scenarios can now lead to a steepenig of the metallicity gradient: (i) the metallicity of the star forming gas in the outskirts of the galaxies is diluted by the freshly accreted gas; (ii) the central parts of the galaxy keep enriching more strongly than the outskirts or (iii) a mix of both.
In fact, as was shown in \citet{Buck2020} in their Fig. 8 scenario (iii) is happening in these galaxies. For most of the time the ISM/cold gas metallicity evolution is self-similar and almost monotonically increasing at each radius but with markedly dilution events that correlate with merger events. In order to investigate this in more depth and study its effect on the metallicity gradient, we plot in Fig.~\ref{fig:feh_evolution} the difference in metallicity of the cold gas, $\Delta\mathrm{[Fe/H]}=\mathrm{[Fe/H]}(R<2\, \mathrm{kpc})-\mathrm{[Fe/H]}(R)$, between the gas in a given annulus at a radius $R$ and the metallicity in the central parts ($R<2$ kpc) of the galaxy, $\mathrm{[Fe/H]}(R<2\, \mathrm{kpc})$ as a function of time. Thus, a flat line in Fig.~\ref{fig:feh_evolution} would indicate perfect self-similar evolution while a more negative value indicates that a certain radius lacks behind the central galaxy in [Fe/H] enrichment and thus a steepening of the metallicity gradient. On the other hand, a progressively more positive or less negative value shows that the radius at hand enriches more strongly in metallicity than the central parts an thus indicates a flattening of the metallicity gradient. In this way, we highlight deviations from the self-similar evolution which is indicating a steepening or flattening of the metallicity gradient as measured over different radial ranges and we are able to disentangle wether its the outskirts or the central parts that dominate the steepening/flattening of the gradient.

Figure~\ref{fig:feh_evolution} complements our findings from Fig.~\ref{fig:half_mass} and shows that during phases of strong steepening of the metallicity gradient (gray vertical lines) it is especially the outskirts of the galaxies (outside of the solar circle, $R>8$ kpc) that lack behind the central parts in [Fe/H] enrichment. In general, we find that the inner annulus of $3<R<5$ kpc (black line) evolves almost self-similar to the central galaxy and relatively unaffected by the mergers. The outer galaxy annuli (blue, orange and red line) show strong negative deviations/dips at the times of mergers (compare also to Fig.~\ref{fig:merger_ratio}) or similar times of strong increase in the cold gas half-mass radius (see Fig.~\ref{fig:half_mass} which indicates that the cold gas [Fe/H] enrichment at these radii is strongly influence by the gas brought in by the merging dwarf galaxies. 
Interestingly, after the merger these radii quickly catch up again in [Fe/H] enrichment by the elevated star formation at the galaxy outskirts \citep[see also Fig. 7 in][]{Buck2020}. 
Furthermore, this figure also shows that the gradual flattening of the metallicity gradient at late times (second half of the age of the Universe) as observed in Fig.~\ref{fig:half_mass} is mostly caused by preferential/elevated enrichment of the galaxy outside the solar circle.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Impact on the age-metallicity relation and the [$\alpha$/Fe] vs. [Fe/H] plane} \label{sec:results-age}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We now turn to investigate the question where the early gas rich mergers appear in the age-metallicity relation and the [$\alpha$/Fe] vs. [Fe/H] plane. To this extent we focus on the galaxy g2.79e12 for which the steepening of the metallicity gradient is strongest.

\subsection{The impact on the age-metallicity relation and the [$\alpha$/Fe] vs. [Fe/H] plane}
\label{sec:g2.79e12}

 Figure~\ref{fig:mdf} shows the metallicity distribution function (MDF) for gas (orange histogram) and stars (blue histogram) of three gas rich satellites that merge with the main galaxy at $2.02$ Gyr and $2.57$ Gyr in comparison to the MDF of gas (gray histogram) and stars (black histogram) in the central galaxy. A similar figure for the [$alpha$/Fe] distribution function is shown in the appendix in Fig.~\ref{fig:mdf_oxygen}. Focusing first on the stellar components (step histograms) we see that the stars contributed by the mergers have on average $\sim0.75$ dex lower metallicity compared to the stars of the main galaxy although at the lowest metallicities ($\lesssim-2.5$) the tails of both distributions overlap. Turning to the gaseous component (filled histograms) we find that (i) the gas and stars of the dwarf galaxies have rather similar metallicity distribution functions where the gas seems to have slightly higher metallicity by about $\sim0.1$ dex. The gaseous MDF of the main galaxy on the other hand shows two peaks separated by about $\sim1$ dex. Those two peaks represent the star forming gas disk at higher metallicities and the gas of the  CGM. Comparing the gas MDF of the merger to the one of the main galaxy we recover the findings from the stellar component; the merger has an MDF that peaks $\sim0.75$ dex below the MDF of the main galaxy's gas broadly overlapping with the CGM peak.

\begin{figure*}
    \script{mdf.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/2.79e12_mdf_gas.pdf}
        \vspace*{-1.75em}
        \caption{
            Metallicity distribution function of the main galaxy and the merging satellite shortly before coalescence for all three merger events. Filled histograms show the gaseous MDF while steps show the MDF for stars. In any case, the gas metallicity of the satellite is $\sim0.5-0.75$ dex lower that the main galaxy's gas metallicity.
        }
        \label{fig:mdf}
    \end{centering}
\end{figure*}

\begin{figure*}
    \script{age_feh.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/2.79e12_age_metallicity_grid.pdf}
        \caption{
            Age-metallicity relation for the stars in the galaxy g2.79e12 (black histogram in the background). Individual panels show with color where the stars contributed by the merging dwarf galaxies reside in age-metallicity space. The panel titles highlight the cosmic time at which the merger happens.
        }
        \label{fig:age_feh}
    \end{centering}
\end{figure*}

\begin{figure*}
    \script{feh_ofe.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/2.79e12_feh_ofe_grid.pdf}
        \caption{
            [$\alpha$/Fe] vs. [Fe/H] plane for the stars in the galaxy g2.79e12 (black histogram in the background). Individual panels show with color where the stars contributed by the merging dwarf galaxies reside in this plane (upper row) and where the gas from the merging dwarf galaxies contributes to star formation after the merger completed (lower panel).
        }
        \label{fig:ofe_feh}
    \end{centering}
\end{figure*}

@@

Recent observations of the MW by \citet{Lu2022b} using data from LAMOST combined with {\it Gaia} eDR3 surprisngly found a steepening of the metallicity gradient at the time of the GSE merger \citep{Belokurov2018, Helmi2018}. In this work we set out to investigate the influence of early massive merger events on the evolution of the metallicity gradient of the cold star forming gas disk of simulated galaxies. We made use of the NIHAO-UHD suite \citep{Buck2020} of cosmological hydrodynamical simulations of MW-mass galaxies to study the frequency of massive early mergers (see Fig.~\ref{fig:merger_ratio}) and their detailed impact on the gaseous disks in the simulated galaxies.

Our results are summarized as follows:
\begin{itemize}
%
\item We find that all four galaxies used in this work exhibit a strong steepening of the metallicity gradient at early times much like the observations of \citet{Lu2022b} revealed for the MW (see blue line in Fig.~\ref{fig:half_mass}).  
%
\item Our simulations show that a steepening of the cold gas metallicity gradient is caused by massive, gas rich early merger events that cause a sudden increase in the cold gas disk size (see orange line in Fig.~\ref{fig:half_mass}).
%
\item The merging dwarf galaxies mostly affect the galaxy outskirts and lead to an increase in cold gas surface density of up to 200\% (see Fig.~\ref{fig:surf_den}).
%
\item In more detail, the addition of relatively un-enriched, fresh gas by the mergers break the self-similar enrichment of the ISM and cause a dilution of the cold gas in the outskirts of the galaxies (see Fig.~\ref{fig:feh_evolution}) which in turn leads to a steepeing of the metallicity gradient.
%
\end{itemize}
